The Iterative Signature Algorithm

### 2009-10-30 isa2/inst/doc/ISA_tutorial.Rnw
Simplify Makefile, reorganize HTML generation from the vignettes.

% \VignetteIndexEntry{The Iterative Signature Algorithm}
\documentclass{article}
\usepackage{ragged2e}

\newcommand{\Rfunction}[1]{\texttt{#1}}
\newcommand{\Rpackage}[1]{\texttt{#1}}
\newcommand{\Rclass}[1]{\texttt{#1}}
\newcommand{\Rargument}[1]{\textsl{#1}}
\newcommand{\filename}[1]{\texttt{#1}}
\newcommand{\listentry}[1]{\texttt{#1}}

@@

<<Roptions,echo=FALSE,results=hide>>=
## Some options for R
options(width=60)
@

@@

To run the typical ISA work flow with default parameters on your data
matrix, just load the \Rpackage{isa2} package and call the
\Rfunction{isa} function with your matrix as the single argument. The
return value of \Rfunction{isa} is a named list, the members
\texttt{rows} and \texttt{columns} contain the biclusters ISA have
found. Every bicluster is given by a pair of columns from
\texttt{rows} and \texttt{columns} (i.e. the first columns define the
first bicluster, etc.) and the elements the biclusters are the
non-zero elements in the columns of \texttt{rows} and
\texttt{columns}.

@@

The Iterative Signature Algorithm (ISA) is biclustering method. Its 
input is a matrix and its output is a set of biclusters: blocks of the
potentially reordered input matrix, that fulfill some predefined
criteria. A biclustering algorithm tries to find
blocks that are different from the rest of the matrix, e.g. the values
covered by the bicluster are all above or below the background.

@@

Fig.~\ref{fig:twoclusters} shows possibly the simplest
example of a (rather artificial) data matrix with very strong modular
structure. It is a $20\times 20$ matrix and (after reordering its rows
and columns) it has two correlated blocks,
each of them of size $10\times 10$.
\begin{figure}
<<toymatrix,echo=FALSE,fig=TRUE,width=6,height=3>>=
library(isa2)
library(Matrix)
data <- data2 <- isa.in.silico(20,20,2,10,10)[[1]]
perm1 <- sample(seq_len(nrow(data)))
data2 <- data2[perm1,]
perm2 <- sample(seq_len(ncol(data)))
data2 <- data2[,perm2]

@@

ISA works in an iterative way. For an $E (m\times n)$
input matrix it starts from a seed vector $r_0$, which is
typically a sparse 0/1 vector of length $m$. The non-zero elements in
the seed vector define a set of rows in $E$. Then the transposed of
$E$, $E'$ is multiplied by $r_0$ and the result is thresholded.

The thresholding is an important step of the ISA, without thresholding
ISA would be equivalent to a (not too effective) numerical singular
value decomposition (SVD) algorithm. Currently thresholding is done by
calculating the mean and standard deviation of the vector and keeping
only elements that are further than a given number of standard
deviations from the mean. Based on the \Rargument{direction} parameter,
this means keeping values that are significantly higher than the
mean (``\emph{up}''), or keeping the ones that are significantly lower
than the mean (``\emph{down}''); or keeping both
(``\emph{updown}'').

@@

It is very possible that the ISA finds the same module more than once;
two or more seeds might converge to the same module. The function
\Rfunction{isa.unique} eliminates every module from the result of 
\Rfunction{isa.iterate} that is very similar (in terms of
Pearson correlation) to the one that was already found before.

It might be also apparent from the description of ISA, that the
biclusters are soft, i.e. they might have an overlap in their rows,
columns, or both. It is also possible that some rows and/or
columns of the input matrix are not found to be part of any ISA
biclusters. Depending on the stringency parameters in the
thresholding (i.e. how far the values should be from the mean), it
might even happen that ISA does not find any biclusters.

@@

Using in silico data we observed that ISA has the best performance if the
input matrix is normalized (see \Rfunction{isa.normalize}). The
normalization produces two matrices: $E_r$ and
$E_c$. $E_r$ is calculated by transposing $E$ and
centering and scaling its rows (see the \Rfunction{scale} R function). $E_c$ is
calculated by centering and scaling the rows of $E$. $E_r$ is
used to calculate the column signature of rows and $E_c$ is used
to calculate the signature of the columns.

@@

In addition to finding biclusters in the input matrix, the ISA also
assigns scores to the rows and columns, separately for each
module. The scores are between minus one and one and they are by
definition zero for the rows/columns that are not included in the
module. For the non-zero entries, the further the score of a
row/columns is from zero, the stronger the association between the
row/column and the module. If the sign of two rows/columns are the
same, then they are correlated, if they have opposite signs, then they
are anti-correlated.

\subsection{Robustness}
As ISA is an unsupervised algorithm, it may very well find some
modules, even if you feed in noise as an input matrix. To avoid these
spurious modules we defined a robustness measure, a single number for
a module that measures how well the rows and the columns are
correlated. The robustness score is a generalization of the first
singular value of the matrix. If there would be no thresholding during
the ISA iteration, then
the ISA would (almost always) converge to the leading SVD vector and
the robustness score would be the corresponding singular value.

@@

Let us see an example. We load the \Rpackage{isa2} package, and
generate some in 
silico data with the \Rfunction{isa.in.silico} function first:
<<insilico,cache=TRUE>>=
library(isa2)
data <- isa.in.silico(200,200,2,50,50)[[1]]
@
The function call above generates a data matrix similar to the one in
Fig.~\ref{fig:twoclusters}, but a bigger one. The first two arguments define the size of
the data matrix ($200\times 200$), the third the number of modules (two)
and the fourth and fifth give the size of the modules, both are
$50\times 50$ submatrices in this case. By default
\Rfunction{isa.in.silico} creates non-overlapping modules with some
background noise.

@@

<<blockplot.pre,echo=FALSE,results=hide>>=
png("blockplot.png", 800, 600)
@ 
<<blockplot>>=
images(list(data), xlab="", ylab="", strip=FALSE)
@ 
<<blockplot.post,echo=FALSE,results=hide>>=
dev.off()
@

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{blockplot}
\caption{Another artificial data matrix with two modules and a lot of
  noisy elements.}
\label{fig:block}
\end{figure}

All we have to do now is calling the \Rfunction{isa} function on the
data matrix. 
<<seed,echo=FALSE>>=
## Just to get the same results
set.seed(10)
@ 
<<isa1,cache=TRUE>>=
modules <- isa(data)
names(modules)
@

The \Rfunction{isa} function returns a list with four
elements. \listentry{rows} and \listentry{columns} define the
modules. The non-zero elements in the first column of both define the
first module, the second columns define the second module, etc.
The number of columns in \listentry{rows} (and \listentry{columns})
correspond to the number of modules.
<<no.mods.>>=
ncol(modules$rows)
@
ISA has found \Sexpr{ncol(modules$rows)} %$
modules in the artificial data. This seem a bit surprising, as we
expected only two modules, so let us take a look at the modules. We
simply plot the row scores of the individual modules
(Fig.~\ref{fig:simpleplot}). The first two modules corresponds to the
two blocks in the input matrix. The third module is the union of them,
with opposite score signs. The \Rfunction{isa} function also finds modules
containing anticorrelated rows/columns, as long as the
anti-correlation is coordinated. I.e. in the third module, the first
50 rows \emph{always} behave in an opposite way compared to the second
50 rows. This is why they are collected into a single module. The
following code creates Fig.~\ref{fig:simpleplot}.

<<simpleplot,results=hide,include=FALSE>>=
layout(cbind(seq(ncol(modules$rows))))
sapply(seq(ncol(modules$rows)), function(x) {
  par(mar=c(3,5,1,1))
  plot(modules$rows[,x], ylim=c(-1,1), 
       ylab="scores", xlab=NA, 
       cex.lab=2, cex.axis=1.5)
})
@

@@

The \listentry{rundata} entry of the return value of \Rfunction{isa}
contains information about the ISA run, all the parameters (the default
values in our case) are included here. It has the following members:
<<isaresult.rundata>>=
names(modules$rundata)
@ 
See the documentation of \Rfunction{isa} for more about these.

@@

In this section we will perform each steps of the ISA analysis
individually. This makes sense only if the user wants to adjust the
parameters of some steps. Otherwise a simple call to the
\Rfunction{isa} function would do.

@@

\begin{figure}
\centering
\includegraphics{tenmods}
\caption{In-silico data with ten non-overlapping modules.}
\label{fig:tenmodules}
\end{figure}

@@

To run the ISA, we need to create the starting seeds first. The
\Rfunction{isa} function uses sparse random seeds, generated by
\Rfunction{generate.seeds}. We also have the choice of using
non-random seeds, e.g. if you have a matrix with gene expression data
measured on many samples you can use ``gene'' seeds that correspond to
biological pathways; in this case ISA does a biased search to find
modules related to the input pathways. Or, in a case/control
experiment one can use seed vectors that correspond to cases and look
for modules that are different in the control and the
case samples.

For now, we will use thousand random seeds, with different sparseness:
<<generate.seeds>>=
row.seeds <- generate.seeds(length=nrow(data[[1]]), count=1000, 
                            sparsity=c(2,5,10,100))
@ 
Note, that vector giving the sparsity is recycled, thus we will have
four kind of seeds, some with 2, 5, 10 and 100 rows; 250 of each of
them.

@@

It is also possibly that some seeds converge to an all-zero vector, we
have \Sexpr{sum(apply(modules$rows==0, 2, all))} such module: % $
<<all.zero.modules>>=
sum(apply(modules$rows==0, 2, all))
@

The \Rfunction{isa.unique} function finds duplicate or very similar
modules in a list returned by \Rfunction{isa.iterate} and keeps only a
single one of them. It also eliminates all-zero modules and seeds
that did not converge, \verb+NA+
columns in the \listentry{rows} and \listentry{columns} matrix
correspond to non-convegent seeds:
<<unique>>=
modules2 <- isa.unique(normed.data, modules, cor.limit=0.9)
@
The \Rargument{cor.limit} argument specifies the correlation limit
above which two modules are considered to be the same. Let us see how
many modules are left:
<<isa.unique.res>>=
ncol(modules2$rows)
@

@@

To address this problem, we have developed a robustness measure, a
single scalar number that gives how correlated the rows and columns of
a given module are. This measure is a simple generalization of the
singular value decomposition (SVD) singular value, and can be
calculated with the \Rfunction{robustness} function:
<<robustness>>=
rob <- robustness(normed.data, modules2$rows, modules2$columns)
summary(rob)
@ 
As the user can see, the robustness of the \Sexpr{ncol(modules2$rows)} %$
modules varies considerably, from \Sexpr{round(min(rob),2)} to
\Sexpr{round(max(rob), 2)}. For more insight we create a boxplot for
the robustness scores (Fig.~\ref{fig:robbox}).

@@

One can use the robustness measure to filter out list of modules, in
the following way. We permute the input matrix and run the same
module finding procedure that we did for the original matrix. Then we
calculate the robustness scores for both the real and the scrambled
modules and eliminate the ``real'' modules having a robustness score
that is less than the score of at least one scrambled module. Note,
that this filtering can be done only if both sets of modules were
found using the same threshold parameters. The procedure is
implemented in the \Rfunction{isa.filter.robust} function:
<<filter.robust>>=
modules3 <- isa.filter.robust(data[[1]], normed.data, modules2, 
                              perms=1, row.seeds=row.seeds)
@
We used the very same row seeds, as for the original matrix; this is
not strictly required. The \Rargument{perms} argument sets the number
of permutations to perform.

@@

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{othermods}
\caption{The input matrix and five modules with relatively low
  robustness scores. As it turns out, each of these modules contains
  two blocks of the input matrix.}
\label{fig:othermods}
\end{figure}

@@

Visualizing overlapping biclusters is a challenging task. We
show simple methods that usually visualize a single bicluster 
at a time. For some of these we will use the \Rpackage{biclust} R
package.

@@

The \Rpackage{biclust} R package implements several biclustering
algorithms in a unified framework. It uses the class \Rclass{Biclust}
to store a set of biclusters. The \Rfunction{isa.biclust} converts ISA
modules to a \Rclass{Biclust} object. This requires the binarization
of the modules, i.e. the ISA scores are lost, they are converted to
zeros and ones:
<<biclust,cache=TRUE>>=
library(biclust)
five.mods <- isa.in.silico(200,200,5,20,20)
modules <- isa(five.mods[[1]],thr.row=2, thr.col=2)
Bc <- isa.biclust(modules)
Bc
@

@@

There are some examples in this document that show how to create image
plots for the modules and potentially also the input data. See
e.g. Figs~\ref{fig:othermods} and
\ref{fig:overlapplot}. These plots use the \Rfunction{plotModules}
function, that directly takes the output of \Rfunction{isa} (and other
functions with the same output format: \Rfunction{isa.iterate},
\Rfunction{isa.unique}, etc. (In fact \Rfunction{plotModules} calls the
\Rfunction{images} function to do its job. \Rfunction{images} takes a
list of matrices, see Fig.~\ref{fig:noise} for an example.)

The \Rfunction{drawHeatmap} function of the \Rpackage{biclust} package
can be also used to draw an image plot. This visualizes a single
bicluster on top of the input matrix. It reorders the rows and columns
of the input matrix to move the block of the bicluster to the top
left corner of the input matrix (Fig.~\ref{fig:heatmap}).
<<drawHeatmap.pre,echo=FALSE,results=hide>>=
png("drawheatmap.png", 400,400)
@ 
<<drawHeatmap>>=
drawHeatmap(five.mods[[1]], Bc, number=1,local=FALSE)
@ 
<<drawHeatmap.post,echo=FALSE,results=hide>>=
dev.off()
@ 
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{drawheatmap}
\caption{Heatmap plot of the reordered input matrix. The rows and
  columns corresponding to the first bicluster are moved to the
  top-left corner of the matrix.}
\label{fig:heatmap}
\end{figure}

@@

<<noiseplot.pre,echo=FALSE,results=hide>>=
png("noise.png", width=800, height=600)
@ 
<<noiseplot>>=
images(lapply(data, "[[", 1), names=as.character(noise), 
       xlab="", ylab="")
@ 
<<noiseplot.post,echo=FALSE,results=hide>>=
dev.off()
@ 
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{noise}
  \caption{Modular data sets with different levels of noise. The data
  sets are generated by adding normally distributed background noise
  with given variance to the checkerboard matrix.}
  \label{fig:noise}
\end{figure}

@@

<<sensbox,include=FALSE,results=hide>>=
boxplot(best, names=noise, xlab="Noise level", ylab="Sensitivity",
        cex.lab=1.5)
@
\begin{figure}
\centering
<<sensbox2,fig=TRUE,echo=FALSE,results=hide,width=10,height=5>>=
<<sensbox>>
@ 
\caption{The sensitivity of the ISA in the function of noise. The
  sensitivity score is calculated by finding the best (in terms of
  Pearson correlation) ISA module for each block in the input matrix.
  ISA performs very well, even in the presence of relatively high
  noise levels.
}
\label{fig:sensbox}
\end{figure}

@@

ISA biclusters might overlap in their rows, columns, or both. To
illustrate this, we create an artificial data set with two overlapping
blocks.

<<set.seed.overlap>>=
set.seed(1)
@ 
<<overlap>>=
two.over <- isa.in.silico(100,100,2,40,40,overlap.row=10)
@

@@

<<overlap.isa>>=
ov.normed <- isa.normalize(two.over[[1]])
ov.seeds <- generate.seeds(count=100, length=100)
ov.modules.1 <- isa.iterate(ov.normed, ov.seeds, 
                            thr.row=0, thr.col=0, direction="up")
ov.modules.1 <- isa.unique(ov.normed, ov.modules.1)
ov.modules.1 <- isa.filter.robust(two.over[[1]], ov.normed, ov.modules.1)
ov.modules.2 <- isa.iterate(ov.normed, ov.seeds, 
                            thr.row=1, thr.col=1, direction="up")
ov.modules.2 <- isa.unique(ov.normed, ov.modules.2)
ov.modules.2 <- isa.filter.robust(two.over[[1]], ov.normed, ov.modules.2)
ncol(ov.modules.1$rows)
ncol(ov.modules.2$rows)
@

@@

<<overlapplot.pre,echo=FALSE,results=hide>>=
png("overlapplot.png", 800, 600)
@ 
<<overlapplot>>=
plotModules(data=two.over[[1]],
            list(rows=cbind(ov.modules.1$rows, ov.modules.2$rows),
                 columns=cbind(ov.modules.1$columns,
                   ov.modules.2$column)),
            names=c("Input matrix", paste("Module", 1:6)))
@
<<overlapplot.post,echo=FALSE,results=hide>>=
dev.off()
@

\begin{figure}
\centering
\includegraphics{overlapplot}
\caption{The input matrix with two overlapping blocks and the ISA
  modules for this data set. The first three modules were found at
  milder threshold parameters, that is why they are bigger. As you can
  see for Modules \#1 and \#2, at milder thresholds there is a higher
  probability that ISA will pick up some incorrect rows/columns.
}
\label{fig:overlapplot}
\end{figure}

@@

@@

@@

>>>>

% \VignetteIndexEntry{The Iterative Signature Algorithm}
\documentclass{article}
\usepackage{ragged2e}
\usepackage{url}

\newcommand{\Rfunction}[1]{\texttt{#1()}}
\newcommand{\Rpackage}[1]{\texttt{#1}}
\newcommand{\Rclass}[1]{\texttt{#1}}
\newcommand{\Rargument}[1]{\textsl{#1}}
\newcommand{\filename}[1]{\texttt{#1}}
\newcommand{\listentry}[1]{`\texttt{#1}'}

\SweaveOpts{prefix.string=plot}

@@

<<Roptions,echo=FALSE,results=hide>>=
## Some options for R
options(width=60)
options(continue=" ")
if (require(Cairo)) { png <- CairoPNG }
@

@@

To run the typical ISA work flow with default parameters on your data
matrix, just load the \Rpackage{isa2} package and call the
\Rfunction{isa} function with your matrix as the single argument. The
return value of \Rfunction{isa} is a named list, the members
\texttt{rows} and \texttt{columns} contain the biclusters ISA have
found. Every bicluster is given by a pair of columns from
\texttt{rows} and \texttt{columns} (i.e. the first columns define the
first bicluster, etc.) and the elements of the biclusters are the
non-zero elements in the columns of \texttt{rows} and
\texttt{columns}.

@@

The Iterative Signature Algorithm (ISA)~\cite{sa,isa,isamod} is
biclustering method. Its input is a matrix and its output is a set of
biclusters: blocks of the potentially reordered input matrix, that
fulfill some predefined criteria. A biclustering algorithm typically
tries to find blocks that are different from the rest of the matrix,
e.g. the values covered by the bicluster are all above or below the
background.

@@

Fig.~\ref{fig:twoclusters} shows possibly the simplest
example of a (rather artificial) data matrix with very strong modular
structure. It is a $20\times 20$ matrix and---after reordering its rows
and columns---it has two correlated blocks,
each of size $10\times 10$.
\begin{figure}
<<toymatrix,echo=FALSE,fig=TRUE,width=6,height=3>>=
library(isa2)
library(Matrix)
data <- data2 <- isa.in.silico(20,20,2,10,10)[[1]]
perm1 <- sample(seq_len(nrow(data)))
data2 <- data2[perm1,]
perm2 <- sample(seq_len(ncol(data)))
data2 <- data2[,perm2]

@@

ISA works in an iterative way. For an $E$ $(m\times n)$
input matrix it starts from a seed vector $r_0$, which is
typically a sparse 0/1 vector of length $m$. The non-zero elements in
the seed vector define a set of rows in $E$. First, the transposed of
$E$, $E'$ is multiplied by $r_0$ and the result is thresholded.

The thresholding is an important step of the ISA, without thresholding
ISA would be equivalent to a (not too effective) numerical singular
value decomposition (SVD) algorithm. Currently thresholding is done by
calculating the mean and standard deviation of the vector and keeping
only elements that are further than a given number of standard
deviations from the mean. Based on the \Rargument{direction} parameter,
this means keeping values that are significantly higher than the
mean (direction=``\emph{up}''), or keeping the ones that are
significantly lower than the mean (direction=``\emph{down}''); or
keeping both (direction=``\emph{updown}'').

@@

The \Rfunction{isa.iterate} function performs the ISA iteration from a
given set of seed vectors.

It is very possible that the ISA finds the same module more than once;
two or more seed vectors might converge to the same module. The function
\Rfunction{isa.unique} eliminates every module from the result of 
\Rfunction{isa.iterate} that is very similar (in terms of
Pearson correlation) to one that was already found from a different
seed.

The \Rfunction{isa} function performs the whole ISA workflow, this
includes running \Rfunction{isa.iterate} and \Rfunction{isa.unique}.

It might be also apparent, that the ISA
biclusters are soft, i.e. they might have an overlap in their rows,
columns, or both. It is also possible that some rows and/or
columns of the input matrix are not found to be part of any
biclusters. Depending on the stringency parameters in the
thresholding (i.e. how far the values should be from the mean), it
might even happen that ISA does not find any biclusters.

@@

Using in silico data we observed that ISA has the best performance if the
input matrix is normalized (see \Rfunction{isa.normalize}). The
normalization produces two matrices: $E_r$ and
$E_c$. $E_r$ is calculated by transposing $E$, then
centering and scaling its rows (see the \Rfunction{scale} R function). $E_c$ is
calculated by centering and scaling the rows of $E$. $E_r$ is
used to calculate the (column) signature of rows and $E_c$ is used
to calculate the (row) signature of the columns.

@@

In addition to finding biclusters in the input matrix, the ISA also
assigns scores to the rows and columns, separately for each
module. The scores are between minus one and one and they are by
definition zero for the rows/columns that are not included in the
module. For the non-zero entries, the further the score of a
row/columns is from zero, the stronger the association between the
row/column and the module (i.e. the other rows/columns of the
module). If the sign of two rows/columns are the 
same, then they are correlated, if they have opposite signs, then they
are anti-correlated.

\subsection{Robustness}
As ISA is an unsupervised algorithm, it may very well find some
modules, even if you feed in noise as the input matrix. To avoid these
spurious modules we defined a robustness measure, a single number for
a module that measures how well the rows and the columns are
correlated. The robustness score is a generalization of the first
singular value of the matrix. If there would be no thresholding during
the ISA iteration, then
the ISA would (almost always) converge to the leading SVD vector and
the robustness score would be the corresponding singular value.

@@

Let us see an example. We load the \Rpackage{isa2}
package~\cite{isapackage}, and generate some in 
silico data with the \Rfunction{isa.in.silico} function first:
<<insilico,cache=TRUE>>=
set.seed(10)
library(isa2)
data <- isa.in.silico(200,200,2,50,50)[[1]]
@
The function call above generates a data matrix similar to the one in
Fig.~\ref{fig:twoclusters}, but bigger. The first two arguments define the size of
the data matrix ($200\times 200$), the third the number of modules (two)
and the fourth and fifth give the size of the modules, both are
$50\times 50$ submatrices in this case. By default
\Rfunction{isa.in.silico} creates non-overlapping modules with some
background noise.

@@

<<blockplot.pre,echo=FALSE,results=hide>>=
png("blockplot.png", 600, 400)
@ 
<<blockplot>>=
images(list(data), xlab="", ylab="", strip=FALSE)
@ 
<<blockplot.post,echo=FALSE,results=hide>>=
dev.off()
@

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{blockplot.png}
\caption{Another artificial data matrix with two modules and a lot of
  noisy elements.}
\label{fig:block}
\end{figure}

All we have to do now is calling the \Rfunction{isa} function on the
data matrix. 
<<isa1,cache=TRUE>>=
modules <- isa(data)
names(modules)
@

The \Rfunction{isa} function returns a list with four
elements. \listentry{rows} and \listentry{columns} define the
modules. The non-zero elements in the first column of both define the
first module, the second columns define the second module, etc.
The number of columns in \listentry{rows} (and \listentry{columns})
correspond to the number of modules.
<<no.mods.>>=
ncol(modules$rows)
@
ISA has found \Sexpr{ncol(modules$rows)} %$
modules in the artificial data. This seems a bit surprising, as we
expected only two, so let us take a closer look. We
simply plot the row scores of the individual modules
(Fig.~\ref{fig:simpleplot}). The first two modules corresponds to the
two blocks in the input matrix. The third module is the union of them,
with opposite score signs. The \Rfunction{isa} function also finds modules
containing anticorrelated rows/columns, as long as the
anti-correlation is coordinated. I.e. in the third module, the first
50 rows \emph{always} behave in an opposite way compared to the second
50 rows. This is why they are collected into a single module. The
following code creates Fig.~\ref{fig:simpleplot}.

<<simpleplot,results=hide,include=FALSE,eval=FALSE>>=
layout(cbind(seq(ncol(modules$rows))))
sapply(seq(ncol(modules$rows)), function(x) {
  par(mar=c(3,5,1,1))
  plot(modules$rows[,x], ylim=c(-1,1), 
       ylab="scores", xlab=NA, 
       cex.lab=2, cex.axis=1.5)
})
@

@@

The \listentry{rundata} entry of the return value of \Rfunction{isa}
contains information about the ISA run, all the parameters (with the default
values in our case) are included here. It has the following members:
<<isaresult.rundata>>=
names(modules$rundata)
@ 
See the documentation of \Rfunction{isa} for more about these.

@@

In this section we will perform each step of the ISA analysis
individually. This makes sense only if the user wants to adjust the
parameters of some steps. Otherwise a simple call to the
\Rfunction{isa} function would do.

@@

\begin{figure}
\centering
\includegraphics{tenmods.png}
\caption{In-silico data with ten non-overlapping modules.}
\label{fig:tenmodules}
\end{figure}

@@

To run the ISA, we need to create some starting seeds. The
\Rfunction{isa} function uses sparse random seeds, produced by
\Rfunction{generate.seeds}. We also have the choice of using
non-random seeds, e.g. if you have a matrix with gene expression data
measured on many samples you can use ``gene'' seeds that correspond to
biological pathways; in this case ISA does a biased search to find
modules related to the input pathways. Or, in a case/control
experiment one can use seed vectors that correspond to cases and look
for modules that are different in the control and the
case samples.

For now, we will use thousand random seeds, with different sparseness:
<<generate.seeds>>=
row.seeds <- generate.seeds(length=nrow(data[[1]]), count=1000, 
                            sparsity=c(2,5,10,100))
@ 
Note, that vector giving the sparsity is recycled, thus we will have
four kind of seeds, with 2, 5, 10 and 100 rows; 250 of each kind.

@@

It is also possible that some seeds converge to an all-zero vector, we
have \Sexpr{sum(apply(modules$rows==0, 2, all))} such modules: % $
<<all.zero.modules>>=
sum(apply(modules$rows==0, 2, all))
@

The \Rfunction{isa.unique} function finds duplicate or very similar
modules in a list returned by \Rfunction{isa.iterate} and keeps only a
single one of them. It also eliminates all-zero modules and seeds
that did not converge; \verb+NA+
columns in the \listentry{rows} and \listentry{columns} matrix
correspond to non-convegent seeds:
<<unique>>=
modules2 <- isa.unique(normed.data, modules, cor.limit=0.9)
@
The \Rargument{cor.limit} argument specifies the correlation limit
above which two modules are considered to be the same. Let us see how
many modules are left:
<<isa.unique.res>>=
ncol(modules2$rows)
@

@@

To address this problem, we have developed a robustness measure, a
single scalar number that gives how correlated the rows and columns of
a given module are. This measure is a simple generalization of the
singular value decomposition (SVD) singular value, and can be
calculated with the \Rfunction{robustness} function:
<<robustness>>=
rob <- robustness(normed.data, modules2$rows, modules2$columns)
summary(rob)
@ 
The robustness of the \Sexpr{ncol(modules2$rows)} %$
modules varies considerably, from \Sexpr{round(min(rob),2)} to
\Sexpr{round(max(rob), 2)}. For more insight we create a boxplot for
the robustness scores (Fig.~\ref{fig:robbox}).

@@

One can use the robustness measure to filter a list of modules, in
the following way. We permute the input matrix and run the same
module finding procedure that we did for the original matrix. Then we
calculate the robustness scores for both the real and the scrambled
modules and eliminate the ``real'' modules having a robustness score
that is less than the score of at least one scrambled module. Note,
that this filtering can be done only if both sets of modules were
found using the same threshold parameters. The procedure is
implemented in the \Rfunction{isa.filter.robust} function:
<<filter.robust>>=
modules3 <- isa.filter.robust(data[[1]], normed.data, modules2, 
                              perms=1, row.seeds=row.seeds)
@
We used the very same row seeds, as for the original matrix; this is
not strictly required. The \Rargument{perms} argument sets the number
of permutations to perform.

@@

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{othermods.png}
\caption{The input matrix and five modules with relatively low
  robustness scores. As it turns out, each of these modules contains
  two blocks of the input matrix.}
\label{fig:othermods}
\end{figure}

It turns out that these are double-modules, each is a union of two
real modules. ISA finds these, since they also have correlated rows
and columns, although the correlation is lower than for the
single-modules. The double-modules are typically found at lower
thresholds.

@@

Visualizing overlapping biclusters is a challenging task. We
show simple methods that usually visualize a single bicluster 
at a time. For some of these we will use the \Rpackage{biclust} R
package~\cite{biclust}.

@@

The \Rpackage{biclust} R package implements several biclustering
algorithms in a unified framework. It uses the class \Rclass{Biclust}
to store a set of biclusters. The \Rfunction{isa.biclust} function
converts ISA modules to a \Rclass{Biclust} object. This requires the
binarization of the modules, i.e. the ISA scores are lost, they are
converted to zeros and ones:
<<biclust,cache=TRUE>>=
library(biclust)
five.mods <- isa.in.silico(200,200,5,20,20)
modules <- isa(five.mods[[1]],thr.row=2, thr.col=2)
Bc <- isa.biclust(modules)
Bc
@

@@

There are some examples in this document that show how to create image
plots for the modules and potentially also the input data. See
e.g. Figs~\ref{fig:othermods} and
\ref{fig:overlapplot}. These plots use the \Rfunction{plotModules}
function, that directly takes the output of \Rfunction{isa} (and other
functions with the same output format: \Rfunction{isa.iterate},
\Rfunction{isa.unique}, etc.) In fact \Rfunction{plotModules} calls the
\Rfunction{images} function to do its job. \Rfunction{images} takes a
list of matrices, see Fig.~\ref{fig:noise} for an example.

The \Rfunction{drawHeatmap} function of the \Rpackage{biclust} package
can be also used to draw an image plot. This visualizes a single
bicluster on top of the input matrix. It reorders the rows and columns
of the input matrix to move the block of the bicluster to the top
left corner of the input matrix (Fig.~\ref{fig:heatmap}).
<<drawHeatmap.pre,echo=FALSE,results=hide>>=
png("drawheatmap.png", 400,400)
@ 
<<drawHeatmap>>=
drawHeatmap(five.mods[[1]], Bc, number=1,local=FALSE)
@ 
<<drawHeatmap.post,echo=FALSE,results=hide>>=
dev.off()
@ 
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{drawheatmap.png}
\caption{Heatmap plot of the reordered input matrix. The rows and
  columns corresponding to the first bicluster are moved to the
  top-left corner of the matrix.}
\label{fig:heatmap}
\end{figure}

@@

<<noiseplot.pre,echo=FALSE,results=hide>>=
png("noise.png", width=600, height=400)
@ 
<<noiseplot>>=
images(lapply(data, "[[", 1), names=as.character(noise), 
       xlab="", ylab="")
@ 
<<noiseplot.post,echo=FALSE,results=hide>>=
dev.off()
@ 
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{noise.png}
  \caption{Modular data sets with different levels of noise. The data
  sets are generated by adding normally distributed background noise
  with given variance to the checkerboard matrix.}
  \label{fig:noise}
\end{figure}

@@

<<sensbox,include=FALSE,results=hide>>=
boxplot(best, names=noise, xlab="Noise level", ylab="Sensitivity",
        cex.lab=1.5)
@
\begin{figure}
\centering
<<sensbox2,fig=TRUE,echo=FALSE,results=hide,width=10,height=5>>=
<<sensbox>>
@ 
\caption{The sensitivity of the ISA in the function of noise. The
  sensitivity score is calculated by finding the best (in terms of
  Pearson correlation) ISA module for each block in the input matrix.
  Each boxplot is contains ten points, one for each module.
  ISA performs very well, even in the presence of relatively high
  noise levels.
}
\label{fig:sensbox}
\end{figure}

@@

ISA biclusters might overlap in their rows, columns, or both. To
illustrate this, we create an artificial data set with two overlapping
blocks.
<<overlap>>=
set.seed(1)
two.over <- isa.in.silico(100,100,2,40,40,overlap.row=10)
@

@@

<<overlap.isa>>=
ov.normed <- isa.normalize(two.over[[1]])
ov.seeds <- generate.seeds(count=100, length=100)
ov.modules.1 <- isa.iterate(ov.normed, ov.seeds, convergence="cor",
                            thr.row=0, thr.col=0, direction="up")
ov.modules.1 <- isa.unique(ov.normed, ov.modules.1)
ov.modules.1 <- isa.filter.robust(two.over[[1]], ov.normed, ov.modules.1)
ov.modules.2 <- isa.iterate(ov.normed, ov.seeds, convergence="cor",
                            thr.row=1, thr.col=1, direction="up")
ov.modules.2 <- isa.unique(ov.normed, ov.modules.2)
ov.modules.2 <- isa.filter.robust(two.over[[1]], ov.normed, ov.modules.2)
ncol(ov.modules.1$rows)
ncol(ov.modules.2$rows)
@

@@

<<overlapplot.pre,echo=FALSE,results=hide>>=
png("overlapplot.png", 600, 400)
@ 
<<overlapplot>>=
no.modules <- ncol(ov.modules.1$rows) + ncol(ov.modules.2$rows)
plotModules(data=two.over[[1]],
            list(rows=cbind(ov.modules.1$rows, ov.modules.2$rows),
                 columns=cbind(ov.modules.1$columns,
                   ov.modules.2$column)),
            names=c("Input matrix", paste("Module", seq_len(no.modules))))
@
<<overlapplot.post,echo=FALSE,results=hide>>=
dev.off()
@

\begin{figure}
\centering
\includegraphics{overlapplot.png}
\caption{The input matrix with two overlapping blocks and the ISA
  modules for this data set. The first three modules were found at
  milder threshold parameters, that is why they are bigger. As you can
  see for Modules \#1 and \#2, at milder thresholds there is a higher
  probability that ISA will pick up some incorrect rows/columns.
}
\label{fig:overlapplot}
\end{figure}

@@

\section{More information}

For more information about the ISA please see the references at the
end of this paper. The ISA homepage is at
\url{http://www.unil.ch/cbg/homepage/software.html}.

For analyzing gene expression data with ISA, we suggest using
BioConductor~\cite{bioc} and the \Rpackage{eisa}~\cite{eisapackage} R
package.

If you want to run ISA on a computer cluster or a multi-processor
machine, then see the vignette titled ``Running ISA in parallel with
the \Rpackage{snow} pacakge'', in the \Rpackage{isa2} package.

@@

\bibitem[Gentleman, 2004]{bioc}
R. Gentleman, V. J. Carey, D. M. Bates, B.Bolstad, M.
Dettling, S. Dudoit, B. Ellis, L. Gautier, Y. Ge, and others (2004).
\newblock Bioconductor: Open software development for computational biology and
  bioinformatics.
\newblock {\em Genome Biology}, Vol. 5, R80

@@

\bibitem[Cs\'ardi, 2009]{eisapackage}
Cs\'ardi, G. (Sep 15, 2009).
\newblock {\em eisa: The Iterative Signature Algorithm for Gene
  Expression Data}.
\newblock R package version 0.1.
